---
title: Premier League Squad Valuations 2019/2020
author: James Stone
date: '2019-08-12'
slug: premier-league-squad-valuations-2019-2020
categories:
  - Data Viz
tags:
  - Football
  - Premier League
  - Scraping
description: ''
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
extrafont::loadfonts(device="win")
```

A new season begins and every team is filled with renewed hope. But this isn't a sport where all competitors are on an equal footing. Using transfermarkt's estimated market value for each player here is the breakdown of total squad value for each Premier League team as the 2019/20 Premier League season kicks off. Will the market value league table at the start of the season reflect the standings after 38 gameweeks?

# Get the data
```{r}
pacman::p_load(tidyverse, rvest, png, grid, RCurl, janitor,
               d3treeR, treemap)
options(scipen = 999)

# colours from EPL website
purple.colour <- "#37003c"
blue.colour <- "#05f0ff"
pink.colour <- "#e90052"

### Get the data from transfermarkt
club_value_html <- read_html("https://www.transfermarkt.co.uk/premier-league/marktwerteverein/wettbewerb/GB1")

club_value_data <- tibble(
  club = club_value_html %>% 
    html_nodes(".no-border-links") %>% 
    html_text(),
  badge = club_value_html %>% 
    html_nodes(".tiny_wappen") %>% 
    html_attr("src"),
  squad_value = club_value_html %>% 
    html_nodes("#yw1 .rechts.hauptlink") %>% 
    html_text() %>% 
    trimws()
)
```

# Clean the data
```{r}
club_value_data <- club_value_data %>% 
  mutate(squad_value_parsed = case_when(
    grepl("bn", squad_value) ~ squad_value %>% 
      str_extract("\\d+\\.\\d+") %>% 
      as.numeric() * 1000000000,
    grepl("m", squad_value) ~ squad_value %>% 
      str_extract("\\d+\\.\\d+") %>% 
      as.numeric() * 1000000
  ))
```

# Plot the values

```{r fig.height=10}
ggplot(club_value_data %>% 
         mutate(over310 = ifelse(squad_value_parsed > 310000000, TRUE, FALSE)) %>% 
         mutate(club = fct_reorder(club, squad_value_parsed)),
       aes(x = club, y = squad_value_parsed)) + 
  geom_bar(stat="identity", 
           colour = blue.colour,
           fill = blue.colour) +
  geom_text(aes(label = ifelse(squad_value_parsed/1000000 > 1000,
                               glue::glue("{club}: £{squad_value_parsed/1000000000}B"),
                               glue::glue("{club}: £{squad_value_parsed/1000000}M")),
                y = ifelse(squad_value_parsed < 310000000,
                           squad_value_parsed + 10000000,
                           10000000),
                colour = over310),
            hjust=0,
            family = "Roboto Condensed") + 
  labs(title = str_wrap("Premier League Squad Value Table 2019/20", 80),
       caption = "Source: transfermarkt.co.uk") +
  coord_flip() + 
  scale_color_manual(values = c('TRUE' = "black", 'FALSE' = "white"), guide="none") + 
  ggthemes::theme_fivethirtyeight() + 
  theme(
    text = element_text(family="Roboto Condensed"),
    axis.text.y = element_blank(),
    axis.text.x = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_rect(fill=purple.colour),
    panel.background = element_rect(fill=purple.colour),
    title = element_text(colour = blue.colour)
  )
```

# Treemap

To further examine these squad values and provide a more detailed summary for readers, let's construct an interactive treemap using team-specific colouring.

## Get the player-level data
```{r}
base_url <- "https://www.transfermarkt.co.uk"
team_urls <- as.character(glue::glue("{base_url}{club_value_html %>% 
  html_nodes('a.vereinprofil_tooltip') %>% 
  html_attr('href') %>% 
  unique()}"))

squad_data <- team_urls %>% purrr::map_dfr(function(u) {
  # delay to be respectful to the host site
  Sys.sleep(runif(1,2,5))
  
  read_html(u) -> html.tmp
  tibble(
    number = html.tmp %>% 
      html_nodes(".rn_nummer") %>% 
      html_text() %>% 
      as.numeric(),
    player = html.tmp %>% 
      html_nodes("span.hide-for-small a") %>% 
      html_text() %>% 
      .[. != ""],
    position = html.tmp %>% 
      html_nodes("#yw1 .inline-table tr+ tr td") %>% 
      html_text(),
    dob = html.tmp %>% 
      html_nodes(".hide+ td") %>% 
      html_text(),
    nationality = html.tmp %>% 
      html_nodes("#yw1 .flaggenrahmen:nth-child(1)") %>% 
      html_attr("title"),
    market_value = html.tmp %>% 
      html_nodes(".rechts.hauptlink") %>% 
      html_text()
  ) %>% 
    mutate(Team = html.tmp %>% html_nodes(".dataName span") %>% html_text())
})
```

## Parse the data and contruct a colour map based on the team colours. 

```{r}
# Parse the numbers to make them the correct class and standardised

squad_data <- squad_data %>% 
  mutate(market_value_parsed = case_when(
    grepl("k", market_value) ~ str_extract(market_value, 
                                           "\\d+") %>% 
      as.numeric() * 1000,
    grepl("m", market_value) ~ str_extract(market_value,
                                             "\\d+") %>% 
      as.numeric() * 1000000,
  )) %>%  
  filter(!is.na(market_value_parsed))

# colour map

colour.map <- tibble(
  Team = squad_data %>% pull(Team) %>% unique(),
  main.col = c("#6caddf", "#dd0000",
               "#020031", "#034694",
               "#e80909", "#ef0107",
               "#274488", "#0101e8",
               "#7f0000", "#ed1c24",
               "#fdbc02", "#ed1a3b",
               "#fbee23", "#000000",
               "#292d6b", "#005daa",
               "#630f33", "#7b003a",
               "#ffee00", "#ec2227")
) %>% 
  left_join(
    squad_data %>% 
      group_by(Team) %>% 
      mutate(n.players = n_distinct(player)) %>% 
      slice(1) %>% 
      ungroup() %>% 
      select(Team, n.players)
  )

col.palette <- character(0)
for (i in 1:nrow(colour.map)) {
  tmp.palette <- colorRampPalette(c(colour.map$main.col[i],"#ffffff"))
  col.palette <- append(col.palette, tmp.palette(colour.map$n.players[i]))
}
```

## Construct the static and interactive treemap using [treemap](https://github.com/mtennekes/treemap) and  [d3treeR](https://github.com/d3treeR/d3treeR)


### Static

```{r}
squad_data %>% 
      group_by(Team) %>% 
      mutate(total_squad_value = sum(market_value_parsed)) %>% 
      ungroup() %>% 
      arrange(-total_squad_value, -market_value_parsed) %>% 
      rownames_to_column("id") %>% 
      mutate(id = fct_reorder(id, 1:n())) -> dat.df

tmplot <- treemap(
    dat.df,
    index=c("Team", "player"),
    vSize="market_value_parsed",
    vColor="id",
    type="categorical",
    position.legend = "none",
    palette = col.palette,
    title = "2019/20 Premier League Squad/Player Values",
    fontfamily.title = "Roboto Condensed",
    fontfamily.labels = "Roboto Condensed",
    align.labels = c("center","center")
  )
```

```{r include=FALSE}
style_widget <- function(hw=NULL, style="", addl_selector="") {
  stopifnot(!is.null(hw), inherits(hw, "htmlwidget"))

  # use current id of htmlwidget if already specified
  elementId <- hw$elementId
  if(is.null(elementId)) {
    # borrow htmlwidgets unique id creator
    elementId <- sprintf(
      'htmlwidget-%s',
      htmlwidgets:::createWidgetId()
    )
    hw$elementId <- elementId
  }

  htmlwidgets::prependContent(
    hw,
    htmltools::tags$style(
      sprintf(
        "#%s %s {%s}",
        elementId,
        addl_selector,
        style
      )
    )
  )
}
```

### Interactive

Note `style_widget()` from [here](https://github.com/d3treeR/d3treeR/issues/10#issuecomment-248098578). Click a team to make them the focus and see a breakdown of individual player value, click the title to zoom out again. 

```{r}
style_widget(
  d3tree3(tmplot,
          rootname = "Premier League Squad Values 2019/20"),
  addl_selector="text",
  style="font-family:Roboto Condensed; font-size:16px;"
)
```